---
author: "Ranieri Dugo"
date: "2026-03-18"
output: 
  html_document:
    code_folding: hide
---

<style>
body {
  font-size: 18pt;  /* increases main text */
}
h1, h2, h3 {
  font-size: 2em;   /* headings */
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Diet Effects on Chicken Weights

## 1 Introduction

We observe weight measurements for 30 chicks over 21 days, denoted by:
$$
y_{i,t},
$$
where $i=1,\dots,30$ indexes the chicks, and $t\in\{0,2,4,\dots,20,21\}$ indexes the day of observation.

```{r, include = TRUE}
library(ggplot2)
library(dplyr)

df <- 
  datasets::ChickWeight |>
  filter(Diet %in% c(1, 2)) |>
  mutate(Diet = recode(Diet, `1` = "Diet 1", `2` = "Diet 2")) |>
  rename(Weight = weight)

df |>
  select(Chick, Diet) |>
  unique() |>
  count(Diet)
```

The chicks were assigned to one of two diets, with 20 chicks following Diet 1 and 10 chicks following Diet 2. 

```{r, include = TRUE}
library(tidyr)
df |>
  select(Chick, Diet, Time) |>
  count(Diet, Time) |>
  pivot_wider(names_from = Diet, values_from = n)
```

Measurements for the chicks following Diet 2 are recorded consistently over the 21 days. Four chicks in Diet 1 have incomplete weight records by day 21. 

We want to determine whether different diets determine different growth in the chicks.

## 2 Exploratory data analysis

### Growth curves

If different diets lead to different growth, we would expect different slopes in the growth curves of chicks.

```{r, include = TRUE}
df |>
  ggplot(aes(x = Time, y = Weight, group = Chick, color = Diet)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  scale_color_manual(
    values = c("Diet 1" = "steelblue", "Diet 2" = "#C44E52"))
```

### Boxplot

Equivalently, one group of chicks would exhibit higher overall weights at the terminal day, $t=21$. 


```{r, include = TRUE}
df |>
  group_by(Chick) |>
  filter(Time == 21) |>
  ggplot(aes(x = Diet, y = Weight, fill = Diet)) +
  geom_boxplot() + 
  theme_minimal() +
  scale_fill_manual(
    values = c("Diet 1" = "steelblue", "Diet 2" = "#C44E52"))
```

While Diet 2 appears to deliver a higher median weight than Diet 1, the difference can be small relative to the variability.

## 3 Simple comparison

### QQ plots

Visually, the final weight distributions under both diets appear close to Gaussian.

```{r, include = TRUE}
df |> 
  filter(Time == 21) |> 
  ggplot(aes(sample = Weight)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "#C44E52") +
  facet_wrap(~ Diet, scales="free") + 
  theme_minimal() +
  labs(
    x = "Theoretical Quantile",
    y = "Empirical Quantile",
    title="QQ Plots of Chick Weight at Day 21")

```

### Test

We can test equality in the means of two Gaussian populations with (possibly) unequal variances using the *Welch's t-test*. 

$H_0$: chicks on different diets have the same weight on $t=21$;

$H_1$: chicks on different diets have different weights on $t=21$.

```{r, include = TRUE}
df |>
  group_by(Chick) |>
  filter(Time == 21) %>%
  t.test(Weight ~ Diet, data = .)
```
  
We fail to reject $H_0$. 

## 4 Linear regression

The previous approach discards information for $t<21$. 

We consider the linear model:

\[
y_{i,t} = \beta_0 + \beta_1 t + \beta_2 \, \mathbf{1}_{\{ \text{Diet}_i = 2 \}} + \varepsilon_{i,t},
\]

where

\[
\mathbf{1}_{\{ \text{Diet}_i = 2 \}} =
\begin{cases}
1 & \text{if chick } i \text{ is on Diet 2}, \\
0 & \text{otherwise}.
\end{cases}
\]

### Estimation and inference

- Ordinary least squares estimation;

- Inference under Gaussian assumption, $\varepsilon_{i,t}\sim\mathcal{N}(0,\Sigma)$.

```{r, include = TRUE}
naive_lm = 
  df %>%
  lm(Weight ~ Time + Diet, data = .)

summary(naive_lm)
```

- The output considers $\Sigma = \sigma^2I$.

### Diagnostics 1: residuals vs fitted values

```{r, include = TRUE}
res_diagnostics_heterosced <- function(model){
  data.frame(
  fitted_values = fitted(model),
  residuals = residuals(model)) |>
  ggplot(aes(x = fitted_values, y = residuals)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#C44E52", linetype = "dashed") +
  labs(x = "Fitted values", y = "Residuals") +
  theme_minimal()
}

res_diagnostics_heterosced(naive_lm)
```

- Linearity of the conditional mean appears adequate; 

- Presence of *heteroscedasticity*: covariates do not capture that variability in weight increases with the weight itself.

### Diagnostics 2: autocorrelation of residuals

Repeated measurements for the same chicks over time induce *autocorrelation* in the residuals.

```{r, include = TRUE}
res_diagnostics_autocorr <- function(model){
  res = residuals(model)
  tmp <- acf(res, plot = FALSE)
  foo <- 1.96 / sqrt(length(res))
  data.frame(Lag = tmp$lag[-1], ACF = tmp$acf[-1]) |>
    ggplot(aes(x = Lag, y = ACF)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_hline(yintercept = c(foo, - foo), linetype = "dashed", color = "#C44E52") +
    labs(x = "Lag", y = "Autocorrelation") +
    theme_minimal()
}

res_diagnostics_autocorr(naive_lm)
```


<!-- # ```{r, include = TRUE} -->
<!-- # plot(naive_lm, 2) -->
<!-- # ``` -->

### Clustered standard errors

- Autocorrelation and heteroscedasticity robust standard errors;

- Block diagonal covariance matrix of residuals;

- $Cov(\varepsilon_{i,t},\varepsilon_{i,s})\ne 0,\forall t,s$ and $Cov(\varepsilon_{i,t},\varepsilon_{j,s})= 0,\forall i\ne j, t,s$.

```{r, include = TRUE}
library(sandwich)
library(lmtest)
cluster_se <- vcovCL(naive_lm, cluster = ~ Chick)
coeftest(naive_lm, vcov = cluster_se)
```

## 5 Mixed-effects model

Explicitly model within-chick variability adding random effects, $b_{0,i}$ and $b_{1,i}$, to the linear model:

\[
y_{i,t} = \beta_0 + \beta_1 t + \beta_2 \, \mathbf{1}_{\{ \text{Diet}_i = 2 \}} 
+ b_{0,i} + b_{1,i} t + \varepsilon_{i,t},
\]

\[
\begin{pmatrix}
b_{0i} \\
b_{1i}
\end{pmatrix}
\sim \mathcal{N}\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma_0^2 & \sigma_{01} \\
\sigma_{01} & \sigma_1^2
\end{pmatrix}
\right), \quad
\varepsilon_{i,t} \sim \mathcal{N}(0, \sigma^2).
\]

Advantages:

- Explicit covariance structure;

- Efficiency gains in estimation.

### Estimation and inference

- Estimation via Restricted Maximum Likelihood;

- Inference for fixed effects using t-statistics and degrees-of-freedom adjustments.

```{r, include = TRUE}
library(lmerTest)
mixed_lm <- lmer(Weight ~ Time + Diet + (Time | Chick), data = df)
summary(mixed_lm)
```

### Diagnostics

```{r, include = TRUE}
res_diagnostics_heterosced(mixed_lm)
res_diagnostics_autocorr(mixed_lm)
```

## 6 Summary

- Chicks gain around 7 grams per day according to both fixed- and mixed-effects models;

- Chicks following Diet 2 exhibit higher average growth, however this effect is not significant;

- According to the mixed-effects model, chicks that are bigger at birth display lower growth over time.

